---
title: Assignment A02
author: Rajneetk
date: '2021-05-26'
slug: assignment-a02
categories: []
tags: []
---


Goodreads Book


Description:

The dataset is from kaggle website about Goodreads book.


1) Dataset information

    Goodreads offers a platform for people interested in talking about books.Analyze book data not only to understand     users reading habits, but also to predict what books are worth reading.

    All books on GoodRead books are included in this data sets.

    The dataset includes 10 columns, including the books and the scores and reviews.


2) The objective is to predict the rating of the book and the analysis which most affects the evaluation.


## Importing data from goodreads.csv file


```{r, warning=FALSE}

library(tidyverse)
library(gridExtra)
library(dplyr)

goodreads <- read_csv("goodreads.csv")
head(goodreads)
names(goodreads)

```



## One table verb
Count observations by group

```{r}

# select(goodreads,Title,`Average Rating`)

goodreads %>%
  select(Title,`Average Rating`) %>%
  count(`Average Rating`) %>%
  arrange(desc(n))
```

## Two table verbs

Given the results, we would sort the predictions with respect to recommend the highest rated books. 

```{r}
library(dplyr)
library(tidyverse)

goodreads %>%
  select(Title:Bookshelves)

goodreads %>%
  select(starts_with("Bookshelves"))

 left_join(`Book Id`,Title)
 
```









