---
title: Assignment A01 based on mtcars dataset 
author: Rajneetk
date: '2021-05-20'
slug: assignment-a01 on mtcars dataset
categories:
  - Data analytics
tags: []
---



Our analysis will start with the data and then show different links between the car's design and the miles it can achieve per gallon. Below is a detailed report with our code and methods.


Description:

The data was extracted from the US magazine Motor Trend 1974 and includes fuel consumption and 10 car design and performance aspects for 32 cars (1973-74 models).



Source Data:

We have used the mtcars dataset integrated in R.

Given 32 observations dataframe on eleven (numeric) variables


* mpg - Miles per gallon
* cyl - Number of cylinders
* disp- Displacement (cu.in.)
* hp  - Gross horsepower
* drat- Rear axle ratio
* wt  - Weight (1000 lbs)
* qsec- 1/4 mile time
* vs  - Engine (0 = V-shaped, 1 = straight)
* am  - Transmission (0 = automatic, 1 = manual)
* gear- Number of forward gears
* carb- Number of carburetors


look at the structure of the mtcars dataset.

```{r,message=FALSE}

str(mtcars)
summary(mtcars)
library(dplyr)
mtcars

```


# Gross Horsepower vs Miles per gallon

The effect of this horsepower number on mpg is shown below. Linear regression is then applied to fit the data in a line. With the linear model method, we use geom smooth.


```{r,message=FALSE}
library(ggplot2)

ggplot(mtcars, aes(x=hp,y=mpg)) +
  geom_point(color="blue") +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title= "Horsepower vs MPG",x="Gross horsepower",y= "Miles per gallon",caption ="Source:Henderson and Velleman (1981), Building multiple regression models interactively. Biometrics, 37, 391–411")
  
```

It shows that as the horsepower is increasing,gallon is unlikely to hit zero.






## Faceting: individual parts for each vactor level

Impact of Cylinders Number on MPG:

We now carry out the same analysis as above but rather consider the number of cylinders and their effect on gallon(mpg).



```{r,message=FALSE,fig.show='hide', echo=FALSE, results='hide'}
library(ggplot2)

mtcars$cyl <- as.factor(mtcars$cyl)
plot(mpg ~ hp , mtcars, 
    col=cyl, pch=c(4,6,8)[mtcars$cyl], cex=1.2)
legend("topright", legend=levels(mtcars$cyl),
                   pch = c(4,6,8),
                   col=levels(mtcars$cyl))



```



```{r,message=FALSE}

library(ggplot2)

plt <-
ggplot(mtcars, aes(x=hp, y=mpg, color=cyl, shape=cyl)) +
    geom_point(size=3) +
   geom_smooth(method="lm", aes(fill=cyl))+ 
  labs(title= "Facet:Horsepower vs MPG",x="Gross horsepower",y= "Miles per gallon",
       caption ="Source:Henderson and Velleman (1981), Building multiple regression models interactively. Biometrics,37, 391–411") 
plt + facet_wrap(~cyl)
```



### Bar plots

A bar charts is an excellent way to show categorical x-axis variables.

```{r, message=FALSE, echo=FALSE}
library(ggplot2)
library(dplyr)

ggplot(mtcars, aes(factor(cyl),
        fill = factor(cyl))) +
    geom_bar()
```







#### Histograms

To get a sense of how certain data is being analyzed, we compare histograms, first with mpg(miles per gallon) and the second with hp(Gross horsepower).


1: Distribution of Cars by Miles per Gallon

```{r, message=FALSE}
library(ggplot2)

ggplot(mtcars, aes(mpg)) +
  geom_histogram(binwidth = 5) + xlab('Miles per Gallon') + ylab('Number of Cars') + 
   ggtitle('Distribution of Cars by Miles per Gallon')


```


2: Distribution of Cars by Gross horsepower

```{r, message=FALSE}
library(ggplot2)

ggplot(mtcars, aes(hp)) +
  geom_histogram(binwidth=10) + xlab('Gross horsepower') + ylab('Number of Cars') +
  ggtitle('Distribution of Cars by Gross horsepower')
```






